---
title: "V03 - ANOVA Analysis on ETo"
author: "David Dorchies"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

In this vignette we try to do an ANOVA analysis on annual ETo an a given cell of 
the climate dataset: 

```{r}
sf_safran_cells <- sf::read_sf("safran_cell.gpkg")
library(tmap)
cell <- 8426
sf_selection <- sf_safran_cells |> filter(cell == .env$cell)
tmap_mode("view")
tm_basemap(server = "OpenStreetMap") +
  tm_shape(sf_safran_cells) +
  tm_borders() +
  tm_text(text = "cell", auto.placement = TRUE) +
  tm_shape(sf_selection) +
  tm_polygons(col = "red", border.col = "red", lwd = 2, alpha = 0.5)
```

This operation needs several steps:

- get the list of climat scenarios to process
- create a function for computing annual ETo
- read and compute annual ETo for all the scenarios
- Run the QUALYPSO method
- Display ANOVA results with time descriptor

## List the available scenarios

Complete list of scenarios is:

```{r climScenarios}
clim_scenarios <- airGRccia::drias_list(safran = FALSE)
```

Up to now we don't have "CFDt" bias correction, so we remove it from the data set:

```{r filter-climScenarios}
clim_scenarios <- clim_scenarios |> 
  filter(BC != "CDFt", scenario != "historical")
str(clim_scenarios)
```

Unfortunately, the downloading of all the scenarios doesn't seem to been 
successfull for all the scenarios, we need to remove the rows on which the 
files are not available.

```{r}
scenario_availability <- sapply(seq(nrow(clim_scenarios)), function(i) {
  airGRccia::climate_is_available(
    RCP = clim_scenarios$scenario[i],
    GCM = clim_scenarios$GCM[i],
    RCM = clim_scenarios$RCM[i],
    BC = clim_scenarios$BC[i]
  )
})
clim_scenarios <- clim_scenarios[scenario_availability, ]
```

The final table of scenarios can be summarize as below:

```{r}
knitr::kable(
  clim_scenarios |> 
    ungroup() |> 
    mutate(X = "X") |> 
    select(X, gcm_rcm, scenario) |> 
    tidyr::pivot_wider(names_from = scenario, values_from = X)
)
```


We have one row by combination of RCP, GCM, RCM which corresponds to two netCDF
files (historical + selected RCP). 
The function `airGRccia::read_climate` takes care of merging both (See vignette V02).

Each NetCDF contains two ETo: `evspsblpot_FAO` and `evspsblpot_Hg0175` for 
respectively FAO formula and Hargrave formulation.

## Function for computing annual 

```{r FnAnnualETo}
calc_annual_ETo <- function(dates, ETo) {
  years <- lubridate::year(dates)
  return(stats::aggregate(ETo, by = list(years), FUN = sum))
}
```


## Read and compute annual ETo for all the scenarios


We loop over the scenarios and populate two tables:
- `scenAvail`: description of each time series by its characteristics: RCP, GCM, 
  RCM, ETo formula
- `Y`: ETo times series with one row by scenario

```{r read-compute-annual-ETo, error = TRUE}
# First, we create an empty data frame for storing characteristics of each scenario:
scenAvail <- data.frame(
  RCP = character(0),
  GCM = character(0),
  RCM = character(0),
  BC = character(0),
  ETo = character(0)
)

# And an empty data.frame for storing all yearly time series 
# rows = scenarios and cols = years
Y <- data.frame()

# Loop over climatic scenarios (mixing RCPs, GCM-RCMs (and BCs if available))
for (i in seq(nrow(clim_scenarios))) {
  RCP <- clim_scenarios$scenario[i]
  GCM <- clim_scenarios$GCM[i]
  RCM <- clim_scenarios$RCM[i]
  BC <- clim_scenarios$BC[i]
  s <- airGRccia::climate_read(GCM, RCM, BC, RCP)
  if (is.logical(s) && s == FALSE) {
    # An error when reading data was detected, we skip that turn
    next
  }
  dates <- stars::st_get_dimension_values(s, "time")
  clim_variables <- names(s)
  s <- airGRccia::climate_extract_cells(cell, s)
  for (formul_ETo in clim_variables[grep("evspsblpot", clim_variables)]) {
    message("Processing ", RCP, " ", GCM, " ", RCM, " ", formul_ETo)
    # Get the corresponding ETo
    ts_ETo <- s[[formul_ETo]]
    # Compute yearly sum
    yearly_ETo <- calc_annual_ETo(dates, ts_ETo)
    # Format the result as a table with one row and year as column names
    df_yearly_ETo <- matrix(yearly_ETo[[2]], nrow = 1)
    if (any(is.na(df_yearly_ETo))) {
      # If NAs are present that means that the data is missing and we skip this turn
      next
    }
    colnames(df_yearly_ETo) <- yearly_ETo[[1]]
    df_yearly_ETo <- as.data.frame(df_yearly_ETo)
    # Bind new result with already stored results (take care of missing years)
    # See ?bind_rows
    Y <- dplyr::bind_rows(Y, df_yearly_ETo)
    scenAvail <- dplyr::bind_rows(scenAvail,
                                  data.frame(
                                    RCP = RCP,
                                    GCM = GCM,
                                    RCM = RCM,
                                    BC = BC,
                                    ETo = sub("evspsblpot_", "", formul_ETo)
                                  ))
  }
}

# Conversion of Y as a matrix for QUALYPSO
Y <- as.matrix(Y)
```

Let's control the resulting data

```{r}
scenAvail
```

```{r}
str(Y)
colnames(Y)
# Columns are not sorted correctly...
Y <- Y[, sort(colnames(Y))]
```


## ANOVA analysis

QUALYPSO doesn't like a single element in one scenario characteristic.
We need to remove the BC column if only one BC is present in the scenarios

```{r}
if (length(unique(scenAvail$BC)) == 1) {
  scenAvail$BC <- NULL
}
```

Let's run QUALYPSO on this data set!

```{r pressure, echo=FALSE}
library(QUALYPSO)
listOption = list(
  typeChangeVariable = 'rel', # relative evolution
  ANOVAmethod = "QUALYPSO"
)
QUALYPSO.synth <- QUALYPSO(
  Y = Y,
  scenAvail = scenAvail,
  X = as.numeric(colnames(Y)),
  listOption = listOption
)
```

And plot the results!

Main RCP effect on evolution of ETo

```{r}
plotQUALYPSOeffect(QUALYPSO.synth, nameEff = "RCP", xlab = "Years", main = "Relative effect of each RCP on ETo")
```

We can plot the total decomposition of variance

```{r}
plotQUALYPSOMeanChangeAndUncertainties(QUALYPSO.synth, xlab = "Years")
```




